---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 6 - Due date 03/16/22"
author: "Ben Joseph"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
---
  
## Directions
  
  You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
#install.packages("sarima")
library(sarima)
```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 

\begin{enumerate}[label=(\alph*)]

\item AR(2)

> Answer: The ACF will decay exponentially with time. The PACF will reduce to near zero after the lag 2 bar.

\item MA(1)

> Answer: The ACF will reduce to near zero after the lag 1 bar. The PACF will decay exponentially with time. 

\end{enumerate}

## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

```{r}
arma10 <- arima.sim(list(order = c(1,0,0), ar = 0.6), n = 100)
arma01 <- arima.sim(list(order = c(0,0,1), ma = 0.9), n = 100)
arma11 <- arima.sim(list(order = c(1,0,1), ar = 0.6, ma = 0.9), n = 100)
arma10
head(arma10, 10)
head(arma01, 10)
head(arma11, 10)
plot(arma10)
plot(arma01)
plot(arma11)
```


\begin{enumerate}[label=(\alph*)]

\item Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r}
par(mfrow=c(1,3))
acf(arma10, type = "correlation", plot=TRUE, main="ACF of ARMA(1,0)", ylim = c(-.5,1))
acf(arma01, type = "correlation", plot=TRUE, main="ACF of ARMA(0,1)", ylim = c(-.5,1))
acf(arma11, type = "correlation", plot=TRUE, main="ACF of ARMA(1,1)", ylim = c(-.5,1))

```


\item Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r}

par(mfrow=c(1,3))
pacf(arma10, plot = TRUE, main="PACF of ARMA(1,0)", ylim = c(-.5,1))
pacf(arma01, plot=TRUE, main="PACF of ARMA(0,1)", ylim = c(-.5,1))
pacf(arma11, plot=TRUE, main="PACF of ARMA(1,1)", ylim = c(-.5,1))

```

\item Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: I would likely be able to identify the ARMA(1,0) and ARMA(0,1) due to their distinctive ACF and PACF characteristics, but it would likely be much harder to identify the ARMA(1,1) because this type of mixed model is much harder to identify at a quick glance.

\item Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: Each time I rerun my code, the values are slightly different, but the value of the lag 1 bar of the ARMA(1,0) PACF tends to be close to the expected phi value of 0.6 but there is often a little bit of error.


\item Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r}
arma2.10 <- arima.sim(list(order = c(1,0,0), ar = 0.6), n = 1000)
arma2.01 <- arima.sim(list(order = c(0,0,1), ma = 0.9), n = 1000)
arma2.11 <- arima.sim(list(order = c(1,0,1), ar = 0.6, ma = 0.9), n = 1000)
plot(arma2.10)
plot(arma2.01)
plot(arma2.11)

par(mfrow=c(1,3))
acf(arma2.10, type = "correlation", plot=TRUE, main="ACF of ARMA(1,0)", ylim = c(-.5,1))
acf(arma2.01, type = "correlation", plot=TRUE, main="ACF of ARMA(0,1)", ylim = c(-.5,1))
acf(arma2.11, type = "correlation", plot=TRUE, main="ACF of ARMA(1,1)", ylim = c(-.5,1))

par(mfrow=c(1,3))
pacf(arma2.10, plot = TRUE, main="PACF of ARMA(1,0)", ylim = c(-.5,1))
pacf(arma2.01, plot=TRUE, main="PACF of ARMA(0,1)", ylim = c(-.5,1))
pacf(arma2.11, plot=TRUE, main="PACF of ARMA(1,1)", ylim = c(-.5,1))

```
`
> Answer: The characteristics in the ACF and PACF are much more pronounced when running the ARMA model with n=1000. The first two would be very easy to identify. The second will be obvious that there is both AR and MA components, but the order will be hard to identify.

> Answer: With a larger number of observations, the lag 1 bar of the ARMA(1,0) PACF varies much less and tends to be much closer to the expected phi value of 0.6 as I rerun the simulation multiple times. We would expect this as the larger the number of observations, the more the randomness of the model cancels itself out, the more the model returns to the expected phi value. 

\end{enumerate}

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

\begin{enumerate}[label=(\alph*)]

\item Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

> p = 1
> d = 0
> q = 1
> P = 1
> D = 0
> Q = 0
> s = 12
> ARIMA(1,0,1)(1,0,0)_12

\item Also from the equation what are the values of the parameters, i.e., model coefficients. 

> phi_1 = 0.7
> phi_12 = 0.25
> theta_1 = 0.1

\end{enumerate}
## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r}
sarima_model <- sim_sarima(model=list(ma=0.5,sar=0.8, nseasons=12), n=1000)
par(mfrow=c(1,3))
plot(sarima_model, type="l", main="SARIMA(0,1)(1,0) Plot")
acf(sarima_model, type = "correlation", plot=TRUE, main="SARIMA(0,1)(1,0) ACF", ylim = c(-.5,1), lag.max = 48)
pacf(sarima_model, plot=TRUE, main="SARIMA(0,1)(1,0) PACF", ylim = c(-.5,1), lag.max = 48)

```

> Answer: I ran the ACF and PACF out to lag 48 to see how the seasonal lag decays over time. Since the seasonal component is an AR of order = 1, we would expect the first seasonal lag (lag=12) in the PACF to equal the seasonal autoregressive coeficient of 0.8 and then all other lags with intervals of 12 will fall off quickly. We would expect all ACF lag multiples of 12 decay over time. 

> Conversely, the non-seasonal trend is a MA model of order = 1. We would expect to see lag 1 in the ACF equal the theta value of .5 and then fall to zero other than lags of multiples of 12. Then we would see the PACF values more slowly decay over time.

> What we observe is that the actual values that we expected to be near but slightly below the expected phi and theta coefficient values. This might be happening because the competing patterns have an effect on each other. Overall, however, the expected patterns are observable.